~\Large Increased sexual dimorphism does not evolve in a fossil stickleback following ecological release from fish piscivores

### 2024-11-25 manuscript/paper.tex
addressing Greg's changes and additions. Some questions back to you.

Briefly, we used multiple imputation (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
of the fossils. Let \(\boldsymbol{W}\) be an
\((n_{extant} + n_{fossil}) \times 1\) vector of the covariate sex of
the stickleback fish and \(\boldsymbol{Y}\) be an
\((n_{extant} + n_{fossil}) \times K\) matrix of the \(K\) phenotypes of
interest. Because the sex of the fossilized stickleback fish is
unobservable, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
observed extant sex and missing fossil sex, respectively. We imputed
missing sex for the fossil data by sampling from the posterior
predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching, implemented in R (CITATION).
Traditionally, the choice for the number of completed dats sets is a
relatively small number such as \(M = 5\) or \(M = 10\). However, Zhou
and Reiter (\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a
larger number of imputed data sets if users intend on performing
Bayesian analysis after imputation. Therefore, we imputed \(M = 100\)
complete datasets.
\textcolor{red}{Greg, PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
\textcolor{red}{Greg, AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}

We then pooled all of the draws from the posterior distribution across
all the imputed data sets to estimate the posterior distributions of our
parameters of interest (i.e., the trait values), rather than using
Rubin's combining rules, following Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}). We then dropped extant
fish from our imputed data sets and used these posterior distributions
in our downstream analyses of fossil sexual dimorphism.

@@

The goal of our analysis was to estimate fossil trait means through
time, by sex, while incorporating uncertainty across imputations in the
classification of sex. As we explain in this section, we are
implementing a complex model structure to account for the fact that the
rate of change in the mean of the trait over time may be a non-constant
value. In addition, we are using multiple imputation procedure to
account for the fact that the sex is unobserved in the fossils in
addition to other pieces of information that are missing. We implement a
Bayesian approach because these methods are more effective for analyzing
hierarchical model structures as well as incorporating missing data.

@@

Continuous traits will likely have some correlation with standard length
(i.e., allometry) (Huxley
(\protect\hyperlink{ref-Huxley1932}{1932}))(Voje, Bell, and Stuart
(\protect\hyperlink{ref-Vojeetal2022}{2022}))(Voje et al.
(\protect\hyperlink{ref-Vojeetal2014}{2014})). We account for this by
adding an additional parameter, \(\gamma_k\). In addition, the
continuous traits ds1, ds2, ds3, lps, tpg are unique in that these
traits tend to be lost over time, results in a proportion of zeros in
our empirical dataset. For these random variables, we add binary
variable \(Z_{ij}\), where \(Z_{k,ij} = 1\) means that the trait is
lost, and \(Z_{k,ij} = 0\) means the trait is observed. For all other
continuous traits, we assume \(Z_{k,ij} = 0\)

If \(Y_{k,ij}\) is a continuous trait, then, potentially conditioning on
\(Z_{k,ij} = 0\), \begin{align}
{Y}_{k,ij} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align} In other words, (\ref{eq:cont}) only considers the subset of
fossils where the given trait is present when performing a data
analysis. For the traits with potential loss, we model the \(Z_{k,ij}\)
via a \(Bernoulli(p_{k,ij})\) distribution where \begin{align}
\text{logit}({p}_{k,ij}) & = \left\{\begin{array}{llll} \alpha_{k,ft_i} + \beta_kY_{K,ij} & W_{ij} = \text{Female} \\ \alpha_{k,mt_i} + \beta_kY_{K,ij}, & W_{ij} = \text{Male} \end{array}\right.,
\label{eq:cont_prob}
\end{align} and logit is a function to convert the probability to the
log-odds, to allow for easier modelling of linear techniques while still
ensuring any probability predictions are guaranteed to be between 0 and
1.

@@

\(\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}\) for trait \(k\)
where \(\tau_r\) is the estimate of the Kendall's \(\tau\) correlation
for \(\mu_{k,m,t_i} - \mu_{k,f,t_i}\) versus time \(t_i\) for the
\(r^{th}\) MCMC sample from the posterior distribution.
\textcolor{red}{Matt, I think there's an absolute value problem here. Dimorphism can increase of M-F gets more positive or more negative. I've indicated that in the text below, but I could use some help refininig the explanation -YS}
For each trait, the proportion of values in \(\boldsymbol{\tau}_k\) that
are greater than 0 (\(\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)\)) may
indicate that males have gotten relatively larger than females,
indicating more pronounced dimorphism. For each trait, the proportion of
values in \(\boldsymbol{\tau}_k\) that are less than 0
(\(\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} < 0)\)) may indicate that
females have gotten relatively larger than males, indicating more
pronounced dimormphism.

@@

The extant data used here consist of a total of 367 specimens all with
known sex (Table S1). Of these, there are 202 and 165 female and male
specimens, respectively.

@@

\textcolor{red}{Greg, WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}

@@

@@

Table \ref{tab:auc-table} shows the probability that the posterior
distribution of the differences in the mean is greater than the
posterior distribution of the differences at the first time point.
Values near 0.5 indicate very little difference between the
distributions whereas values closer to 0 and 1 indicate changes in the
distributions with values of 1 meaning the difference in means has
shifted towards males having a larger mean and values of 0 indicating
the difference in means has shifted towards larger means in females.
These results are shown graphically in figure \ref{fig:AUC-plot}.

@@

@@

@@

>>>>

Fossil and extant data were combined to create a single data set which
contained small amounts on missingness on the observed traits with the
exception of sex, which was completely missing for fossil specimens as
this was unobservable. The sex of the fossil data was considered to be
missing data and was imputed in the same step of the rest of the missing
data using multiple imputation (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})). Imputation was
performed using the multiple imputation by chained equations (MICE)
algorithm (Buuren and Groothuis-Oudshoorn
(\protect\hyperlink{ref-MICE}{2011})) via predictive mean matching (PMM)
(Andridge and Little (\protect\hyperlink{ref-hotdeck}{2010})) and
implemented using the R programming language (R Core Team
(\protect\hyperlink{ref-R2022language}{2022})). Traditionally, the
choice for the number of completed data sets is a relatively small
number such as \(M = 5\) or \(M = 10\). However, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a larger number
of imputed data sets if users intend on performing Bayesian analysis, as
was performed here, after imputation. Therefore, we imputed \(M = 100\)
complete data sets. Extant fish records were then dropped from each of
the imputed data sets, and only the fossil data was used in downstream
analyses of fossil sexual dimorphism.

After performing the Bayesian analysis on each of the \(M = 100\)
imputed data sets, draws from the posterior distribution based on each
of the imputed data sets were pooled together to combine posterior draws
across imputations as recommended in Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) rather than using Rubin's
traditional combining rules.

\textcolor{red}{Greg, PRESUMABLY this also imputes missing data in the other traits? We should probably say that explicitly. -YS}
\textcolor{orange}{Updated - GReg}
\textcolor{red}{Greg, AKHIL'S METHODS FOR VALIDATION OF THE IMPUTATION MODEL GOES HERE}
\textcolor{orange}{Are we sure we want that here?  I was going to add it in later}

@@

The goal of our analysis was to model fossil trait means through time,
by sex, while incorporating uncertainty across imputations in the
classification of sex. We therefore implemented a Bayesian hierarchical
model structure (described below) to account for classfication
uncertainty and the fact that the rate of change of any one trait mean
over time may be a non-constant value.

@@

Continuous traits will likely have some correlation with standard length
(i.e., allometry) (Huxley
(\protect\hyperlink{ref-Huxley1932}{1932}))(Voje, Bell, and Stuart
(\protect\hyperlink{ref-Vojeetal2022}{2022}))(Voje et al.
(\protect\hyperlink{ref-Vojeetal2014}{2014})). We account for this by
adding an additional parameter, \(\gamma_k\). In addition, the
continuous traits ds1, ds2, ds3, lps, tpg are notable because they tend
to be lost during the evolution of armor reduction in this system (Bell,
Baumgartner, and Oslon (\protect\hyperlink{ref-Belletal1985}{1985}),
Stuart, Travis, and Bell
(\protect\hyperlink{ref-Stuartetal2020}{2020})). This generates
meaningful zeros for trait lengths in our empirical dataset that violate
model assumptions. Therefore, for these random variables, we added the
binary variable \(Z_{ij}\), where \(Z_{k,ij} = 1\) means that the trait
is lost, and \(Z_{k,ij} = 0\) means the trait is observed. For all other
continuous traits, we assume \(Z_{k,ij} = 0\)

If \(Y_{k,ij}\) is a continuous trait, then, potentially conditioning on
\(Z_{k,ij} = 0\), \begin{align}
{Y}_{k,ij} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align} In other words, (\ref{eq:cont}) only considers the subset of
fossils where the given trait is present when performing a data
analysis. For the traits with potential loss, we model the \(Z_{k,ij}\)
via a \(Bernoulli(p_{k,ij})\) distribution where \begin{align}
\text{logit}({p}_{k,ij}) & = \left\{\begin{array}{llll} \alpha_{k,ft_i} + \beta_kY_{K,ij} & W_{ij} = \text{Female} \\ \alpha_{k,mt_i} + \beta_kY_{K,ij}, & W_{ij} = \text{Male} \end{array}\right.,
\label{eq:cont_prob}
\end{align} and logit is a function to convert the probability to the
log-odds, to allow for easier modelling of linear techniques while still
ensuring any probability predictions are guaranteed to be between 0 and
1. Modeling traits this way allows us to first estimate the probability
of trait loss for an individual at a given time, and ask if the sexes
differ in their tendency toward loss through time. Then, for individuals
that have not lost the trait, we can estimate trait values and calculate
sexual dimorphism through time.

@@

\(\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}\) for trait \(k\)
where \(\tau_r\) is the estimate of the Kendall's \(\tau\) correlation
for \(\mu_{k,m,t_i} - \mu_{k,f,t_i}\) versus time \(t_i\) for the
\(r^{th}\) MCMC sample from the posterior distribution. For each trait,
a large proportion of values in \(\boldsymbol{\tau}_k\) that are greater
than 0 (\(\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)\)) may indicate
that males have gotten relatively larger than females through time,
indicating an increase in male-biased dimorphism through time. A large
proportion of values in \(\boldsymbol{\tau}_k\) that are less than 0
(\(\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} < 0)\)) may indicate that
females have gotten relatively larger than males, indicating an increase
in female-biased dimorphism through time.

@@

@@

\hypertarget{validating-imputation-of-sex}{%
\subsubsection{Validating Imputation of
Sex}\label{validating-imputation-of-sex}}

{\centering \includegraphics{paper_files/figure-latex/roc-1}

\caption{Reciver-Operating Characteristic Curve comparing imputed sex to actual sex for the extant cases using 5-fold cross validation (AUC = 0.831)}\label{fig:roc}
\end{figure}

In order to validate the efficacy of imputing sex in the fossil data, we
performed 5-fold cross validation using only the extant cases. For a
given fold, we took the true, observed values of sex and removed them to
create missing data similar to what we observe with the fossil data.
Then with one fold held out, we performed multiple imputation with \(M\)
= 5 (using the same model that we use in the actual imputation
(i.e.~MICE with PMM)) using the remaining folds with sex observed to
build a model used to imputed the sex of each fish in the held out fold
(along with any other variables that needed to be imputed). We can then
compare the imputed values of sex to the true values of sex to evaluate
the imputation accuracy of the imputation model. These imputations on
the hold out samples were then combined to estimate the probability of
male vs female (i.e.~the probability of being male was estimated as the
proportion of imputations that predicted male. Using these
probabilities, an ROC curve (Figure \ref{fig:roc}) was constructed to
evaluate the accuracy of the imputations. The AUC was 0.831, indicating
excellent discrimination (Hosmer, Lemeshow, and Sturdivant
(\protect\hyperlink{ref-HosmerLemeshow}{2013}), Page 177) between a fish
being male or female.

\hypertarget{summary-of-imputed-data}{%
\subsubsection{Summary of Imputed Data}\label{summary-of-imputed-data}}

{\centering \includegraphics{paper_files/figure-latex/est-p-1}

\caption{Density estimate of the percentage of specimen that were imputted to be male across the M = 100 imputations}\label{fig:est-p}
\end{figure}

Across the 100 imputations of the 814 fossil specimen, the average
percentage of males in an imputation was 46.64\% with a standard
deviation across imputations of 6.79\% ranging from 30.1\% to 62.29\%.
In only 31 of the 100 imputations were there more males than females
with the remaining 69 imputations having exactly the same number or more
females. Figure \ref{fig:est-p} shows the distribution of the proportion
of males across the \(M\) = 100 imputations across all time periods.

{\centering \includegraphics{paper_files/figure-latex/male-prop-1}

\caption{Density estimates of the distribution of the proportions of male fish at each time period across 100 imputations}\label{fig:male-prop}
\end{figure}

Figure \ref{fig:male-prop} shows density estimates of the distribution
of the proportions of male fish at each time period across 100
imputations along with the median at each time point. We can see that
the first three time points have proportions of males centered around
about 25\% followed by a shift up in proportion of males between time
periods 5 and 9 with median values ranging from 44\% to 66.7\%. Time
periods 10 through 13 show a shift back towards more females with the
median proportions of males ranging from 30.3\% to 41.5\%. The last
several time points, from 14 through 18, all have median male
proportions just above 50\% with the exception of time period 17 with a
median male proportion of 39.3\%.

\textcolor{red}{Yoel, I think I remember you talking about a reason why it made more sense that there were fewer males in the sample tham the females, but I don't remember the details.  Am I remembering this correctly?  Would be nice to add something here about that and say why these imputations are consistent with that.  
-GJM}

@@

We estimated the change in sexual dimorphism over time by calculating
the distribution of Kendall's Tau for each trait. Here, positive values
of Tau indicate that sexual dimorphism (defined as male means minus
female means) increased through time. Average Tau values ranged from X
to Y across traits and the proportion of positive area under the curve
ranged from Z to W (Table Tau). Distributions are available for each
trait in the Supplementary Material (Figures SX-SX+15). In Figure (Tau),
we show representative distributions. For example,

Something about the probability of dimorphim in the probability of loss?
I.e., the prob\_zero plots. For example, ds2\_prob\_zero has lots of
positive area under the curve indicating that the rate of loss in males
and females diverged, such that males were more likely to lose ds2
through time. Interestingly, for the individuals that kept their spines,
dimorphism increased, with females getting larger than males (or males
getting smaller than females).

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-hotdeck}{}}%
Andridge, R. R., and R. J. A. Little. 2010. {``A Review of Hot Deck
Imputation for Survey Non-Response.''} \emph{International Statistical
Review} 78 (1): 40--64.

@@

\leavevmode\vadjust pre{\hypertarget{ref-HosmerLemeshow}{}}%
Hosmer, D. W., Jr., S. Lemeshow, and R. X. Sturdivant. 2013.
\emph{Applied Logistic Regression, 3rd Edition}. Wiley.

@@

\leavevmode\vadjust pre{\hypertarget{ref-R2022language}{}}%
R Core Team. 2022. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.
